# The Breeding Phenology of Boreal Birds

```{r}
library(tidyverse)
library(here)
```

## Data Downloads

Data from: Shifts in timing and duration of breeding for 73 boreal bird species over four decades

```{r}
# This stopped working for some reason, maybe because I already have it, but it says the URL is invalid
#download.file("https://dryad-assetstore-merritt-west.s3.us-west-2.amazonaws.com/ark%3A/13030/m5867r6c%7C1%7Cproducer/73_species.csv?X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Credential=AKIA2KERHV5E3OITXZXC%2F20231019%2Fus-west-2%2Fs3%2Faws4_request&X-Amz-Date=20231019T213852Z&X-Amz-Expires=86400&X-Amz-SignedHeaders=host&X-Amz-Signature=277d5265a2286c5d38f2863a2fe1437363173d1fa0c762ff6d4b7a8fb45adacc",
             # here("data/73_species.csv"))
```

```{r}
bor_nestlings <- read_csv(here("data/73_species.csv"))

bor_traits <- read_csv(here("data/Traits_73_species.csv"))

avonet <- readxl::read_excel(here("data/AVONET Supplementary dataset 1.xlsx"), 
                             sheet = "AVONET1_BirdLife")
```

## Are nestlings showing up earlier in the year over time?

```{r}
all_birds_trend <- bor_nestlings %>% 
  group_by(Year) %>% 
  summarize(mean_doy = mean(Dayofyear))

ggplot(all_birds_trend, aes(Year, mean_doy)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## Breakdown by species

```{r}
species_trends <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(mean_doy = mean(Dayofyear),
            .groups = "drop")

ggplot(species_trends, aes(Year, mean_doy, color = Species)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

Looking pretty useless for data presentation

How about only 5 species?

```{r}
data_richness <- bor_nestlings %>% 
  count(Species)

most_rich <- data_richness %>% 
  arrange(desc(n)) %>% 
  slice(1:5)

most_rich_trends <- bor_nestlings %>% 
  filter(Species %in% most_rich$Species) %>% 
  group_by(Species, Year) %>% 
  summarize(mean_doy = mean(Dayofyear), 
            .groups = "drop")
?aes
ggplot(most_rich_trends, aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

A general pattern, with one conflicting example. What species have the strongest trends?

```{r}
# Finds the slope of the relationship between y and x
trend <- function(x, y) {
  xy_lm <- lm(y ~ x)
  coef(xy_lm)[2]
}

# Calculate the trend for all species
bor_by_year <- bor_nestlings %>% 
  group_by(Year, Species) %>% 
  summarize(BirdByYear = n()) %>% 
  ungroup()
bor_by_year

bor_trends <- species_trends %>% 
  group_by(Species) %>% 
  summarize(doy_trend = trend(Year, mean_doy))

soi <- c("ARDCIN", "LARMIN")
species_trends %>% 
  filter(Species %in% soi) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm")


```

Find the species with the most positive and most negative trends, then plot just those.

```{r}
nrow_bor_trends <- nrow(bor_trends) # Use this later
bor_extreme <- bor_trends %>% 
  arrange(desc(doy_trend)) %>%   # Sort by the day of year trend
  slice(c(1,nrow_bor_trends))   # Keep just the first (most negative trend) and last (most positive trend) rows
bor_extreme

?`%in%`
# Now plot them
bor_extreme2 <- c("RIPRIP", "ARDCIN")
species_trends %>% 
  filter(Species %in% bor_extreme$Species) %>% 
  ggplot(aes(Year, mean_doy, color = Species)) + 
    geom_point() +
    geom_smooth(method = "lm")
```

## Adding in AVONET

```{r}
# First, select and rename important columns
avonet_size <- select(avonet, SciName = Species1, Mass_g = Mass)
bor_sciname <- select(bor_traits, 
                      Species = Abbreviation, 
                      SciName = "Scientific name")


bor_trends_size <- bor_trends %>% 
  left_join(bor_sciname, by = "Species") %>% 
  left_join(avonet_size, by = "SciName") %>% 
  mutate(abs_trend = abs(doy_trend))

?left_join
# Plot it
ggplot(bor_trends_size, aes(Mass_g, abs_trend)) +
  geom_point() +
  geom_smooth(method = "lm")
```
